Tesseroids: forward modeling gravitational fields in spherical coordinates

### 2015-03-30 manuscript.tex
Cleaned up comments from the source

Satellite missions dedicated to measuring the Earth's gravity field
(like CHAMP, GRACE, and GOCE)
have provided geophysicists with almost uniform and global data coverage.
These new data have enabled interpretations on regional and global scales
\citep[e.g.][]{Reguzzoni2013,Braitenberg2015}.
Modeling at such scales requires taking into account the curvature of the
Earth and calculating gravity gradients as well as the traditional
gravitational acceleration.
A common approach to achieve this is
to discretize the Earth into tesseroids (Figure~\ref{fig:tesseroid})
instead of rectangular prisms.
However, the integral formula for the gravitational effects of a tesseroid
must be solved numerically.
Approaches to this numerical integration include
Taylor series expansion \citep{Heck2007, Grombein2013}
and the Gauss-Legendre Quadrature \citep{Asgharzadeh2007}.
Taylor series expansion produces accurate results at low latitudes but
presents a decrease in accuracy towards the polar regions.
This is attributed to tesseroids degenerating into an approximately triangular
shape at the poles.
The Gauss-Legendre Quadrature (GLQ) integration consists of approximating
the volume integral by a weighted sum of the effect of point masses.
An advantage of the GLQ approach is that it can be
controlled by the number of point masses used.
The larger the number of point masses,
the better the accuracy of GLQ integration.
A disadvantage is the increased computation time as the number of
point masses increases.
Thus, there is a trade-off between accuracy and computation time.
This is a common theme in numerical methods.
\citet{Wild-Pfeiffer2008} investigated the use of different mass elements,
including tesseroids, to compute the gravitational effects of topographic
masses.
The author concludes that using tesseroids with GLQ integration gives the best
results for near-zone computations.
However, the question of how to determine the optimal parameters for GLQ
integration remained open.

Previous work by \citet{Ku1977} investigated the use of the GLQ
in gravity forward modeling.
\citet{Ku1977} numerically integrated the vertical component of the
gravitational acceleration of right rectangular prisms.
The author suggested that the accuracy of the GLQ integration depends on
the ratio between distance to the computation point and the distance between
adjacent point masses.
Based on this, \citet{Ku1977} proposed an empirical criterion that the distance
% * <berg@units.it> 2015-03-19T13:46:11.161Z:
%
%  an criterion or proposed empirical criteria
%
% ^ <leouieda@gmail.com> 2015-03-20T17:41:12.218Z:
%
%  It was an criterion
%
between adjacent point masses should be less than
the distance to the computation point.
\citet{Asgharzadeh2007} used  this criteria for the GLQ integration of
the gravity gradient tensor of tesseroids.
To our knowledge, an analysis of how well this ad hoc criteria of
\citet{Ku1977} works for gravity gradient components or for tesseroids
has never been done before.
There has also been no attempt to quantify the error committed in the GLQ
integration when applying the criteria of \citet{Ku1977}.

\citet{Li2011} devised an algorithm to automatically enforce the criteria of
\citet{Ku1977}.
Their algorithm divides the tesseroid into smaller ones instead of increasing
the number of point masses per tesseroid.
% * <berg@units.it> 2015-03-19T13:48:33.551Z:
%
%  sentence is repeated later.
%
% ^ <leouieda@gmail.com> 2015-03-20T17:40:17.120Z:
%
%  I removed it.
%
A tesseroid is divided if the minimum distance to the computation point
is smaller than the largest dimension of the tesseroid.
This division is repeated recursively until all tesseroids obey the criterion.
Then, GLQ integration is performed for each of the smaller tesseroids
using the specified number of point masses.
The advantage of this adaptive discretization over
increasing the number of points masses is that the
total distribution of point masses will be greater
only close to the computation point.
This makes the adaptive discretization more computationally efficient.

\citet{Grombein2013} developed optimized formula for the gravitational fields
of tesseroids using Cartesian integral kernels.
These formulas are faster to compute and do not have singularities at the poles
like their spherical counterparts.
The Cartesian formulae are numerically integrated using a Taylor series
% * <berg@units.it> 2015-03-19T13:49:40.239Z:
%
%  singular or plural?
%
% ^ <leouieda@gmail.com> 2015-03-20T17:41:54.790Z:
%
%  It should be the plural (formulae)
%
expansion as per \citet{Heck2007}.
\citet{Grombein2013} use a near-zone separation to mitigate the increased error
at high latitudes.
In the so called ``near-zone'' of the computation point they use a finer
discretization composed by smaller tesseroids.
This is accomplished by dividing the tesseroids along their horizontal
dimensions.
However, the determination of an optimal size of the near-zone remains an
open question \citep{Grombein2013}.

@@

\noindent
where $\alpha, \beta \in \{x, y, z\}$,
$\rho$ is the density,
$G = 6.673\times10^{-11}\ m^3kg^{-1}s^{-1}$ is the gravitational constant,
and

@@

Below, we describe the algorithm to calculate
the effect of a single tesseroid from the input model
on a single computation point.
The algorithm starts by creating an empty stack of tesseroids.
Then, the stack is initialized with the single input tesseroid.
% * <berg@units.it> 2015-03-19T21:04:01.768Z:
%
%  the description of the operation of stack is not clear- is it the input tesseroid array?
%
% ^ <leouieda@gmail.com> 2015-03-20T18:02:14.018Z:
%
%  I modified the description above and added more text below as well. Is it clearer now?
%
The initialization is done by copying the tesseroid into the stack
and setting \emph{stktop} to zero (the first element).
It is important to note that the stack is not the input tesseroid model.
Instead, it is a buffer used to temporarily store
each stage of the discretization algorithm.

@@

\begin{enumerate}
    \item ``Pop'' the stack (i.e., take the last tesseroid from it).
        This will cause \emph{stktop} to be reduced by one.
        This tesseroid is the one that will be evaluated in the following
        steps.
    \item Compute the distance $d$ (equation~\ref{eq:distance}) between
        the geometric center of the tesseroid and
        the computation point.
    \item Compute the dimensions of the tesseroid $L_\lambda$, $L_\phi$,
        and $L_r$ using equations~\ref{eq:sizelon}-\ref{eq:sizer}.
    \item Check the condition in equation~\ref{eq:condition} for each
        dimension of the tesseroid.
    \item If all dimensions hold the inequality \ref{eq:condition},
        the tesseroid is not divided and its
        gravitational effect is computed using the
        Gauss-Legendre Quadrature
        (equations~\ref{eq:tesspot}-\ref{eq:tesstensor} and~\ref{eq:glq3d}).
        We use a GLQ order of two for all three dimensions
        ($2 \times 2 \times 2 = 8$ point masses)
        by default.
        This value can be changed using
        a command-line argument of the modeling programs.
    \item If any of the dimensions fail the condition:
    \begin{enumerate}
        \item Divide the tesseroid in half along each dimension that failed
             the condition.
        \item Check if there is room in the stack
            for the new tesseroids
            (i.e.,the number of new elements plus \emph{stktop}
             is smaller than the maximum stack size).
             If there isn't,
             warn the user of a ``stack overflow''
             and compute the effect of the larger tesseroid (step 5).
             If there is room in the stack,
             place the smaller tesseroids into
             the stack.
    \end{enumerate}
    \item Repeat the above steps until the stack is empty
        (\emph{stktop} is equal to -1).
\end{enumerate}

@@

The IPython notebooks
and the data generated for the error analysis,
as well as instructions for installing the software
and running the programs,
are also included in
the source code archive that accompanies this article.
Alternatively,
all accompanying material is available
in online repositories\footnote{
https://github.com/pinga-lab/paper-tesseroids}\footnote{
http://dx.doi.org/10.5281/zenodo.16032}.

@@

For the potential $V$, a distance-size ratio $D=1$ guarantees that the curves
for all experiments are below the $0.1\%$ error threshold.
For $g_z$, the same is achieved with $D=1.5$.
Conversely, $g_{zz}$ requires a value of $D=8$ to achieve an error level of
$0.1\%$.
For a computation height of 260 km, the error curve for $g_{zz}$
intercepts the error threshold line at $D=2.5$.
This behavior suggests that the error curves for $g_{zz}$ might depend on the
computation height.
To test this hypothesis, we computed the error curves for $g_{zz}$ at
heights 2, 10, 50, 150, and 260 km.
Figure~\ref{fig:gzz-with-height} shows the results for $g_{zz}$ at varying
computation heights.
Notice that the distance-size ratio required to achieve $0.1\%$ accuracy
decreases as the computation height increases.
% * <berg@units.it> 2015-03-19T21:07:33.445Z:
%
%  what is the height dependence due to? Should the distance not include height already? Is the height a second control parameter, next to the distance ratio? What is that due to? Is it only for gzz or also for the other gradient components? What about the gx?
%
% ^ <leouieda@gmail.com> 2015-03-20T18:23:53.901Z:
%
%  The height is included in the distance and is not a control variable. This happens for all tensor components but not to the potential, gx, gy, and gz. I think this is because the decay factor for the tensor components is 1/r^3 whereas the discreatization algorithm compares D to r/size. So as the  computation point becomes closer. the field increases more rapidly than the algorithm increases the  amount of discretization. That is why you need higher D (more discretization) when you are closer to the tesseroid.
%
For example, computation at 260 km height requires $D=2.5$ whereas at 10 km
height a value of $D=5.5$ is required to achieve the same accuracy.
One can take advantage of this behavior to reduce the distance-size ratio for
computations of the gravity gradient tensor at high altitudes,
saving computation time.

@@

% * <leouieda@gmail.com> 2015-03-20T18:43:18.571Z:
%
%  I have modified the above paragraph to try to explain why the ratio for gzz varies with height.
%

>>>>

Satellite missions dedicated to measuring the Earth's gravity field
(like CHAMP, GRACE, and GOCE)
have provided geophysicists with almost uniform and global data coverage.
These new data have enabled interpretations on regional and global scales
\citep[e.g.][]{Reguzzoni2013,Braitenberg2015}.
Modeling at such scales requires taking into account the curvature of the
Earth and calculating gravity gradients as well as the traditional
gravitational acceleration.
A common approach to achieve this is
to discretize the Earth into tesseroids (Figure~\ref{fig:tesseroid})
instead of rectangular prisms.
However, the integral formula for the gravitational effects of a tesseroid
must be solved numerically.
Approaches to this numerical integration include
Taylor series expansion \citep{Heck2007, Grombein2013}
and the Gauss-Legendre Quadrature \citep{Asgharzadeh2007}.
Taylor series expansion produces accurate results at low latitudes but
presents a decrease in accuracy towards the polar regions.
This is attributed to tesseroids degenerating into an approximately triangular
shape at the poles.
The Gauss-Legendre Quadrature (GLQ) integration consists in approximating
the volume integral by a weighted sum of the effect of point masses.
An advantage of the GLQ approach is that it can be
controlled by the number of point masses used.
The larger the number of point masses,
the better the accuracy of GLQ integration.
A disadvantage is the increased computation time as the number of
point masses increases.
Thus, there is a trade-off between accuracy and computation time.
This is a common theme in numerical methods.
\citet{Wild-Pfeiffer2008} investigated the use of different mass elements,
including tesseroids, to compute the gravitational effects of topographic
masses.
The author concludes that using tesseroids with GLQ integration gives the best
results for near-zone computations.
However, the question of how to determine the optimal parameters for GLQ
integration remained open.

Previous work by \citet{Ku1977} investigated the use of the GLQ
in gravity forward modeling.
\citet{Ku1977} numerically integrated the vertical component of the
gravitational acceleration of right rectangular prisms.
The author suggested that the accuracy of the GLQ integration depends on
the ratio between distance to the computation point and the distance between
adjacent point masses.
Based on this, \citet{Ku1977} proposed an empirical criterion that
the distance between point masses should be greater than
the distance to the computation point.
\citet{Asgharzadeh2007} used  this criterion for the GLQ integration of
the gravity gradient tensor of tesseroids.
To our knowledge, an analysis of how well this ad hoc criteria of
\citet{Ku1977} works for gravity gradient components or for tesseroids
has never been done before.
There has also been no attempt to quantify the error committed in the GLQ
integration when applying the criteria of \citet{Ku1977}.

\citet{Li2011} devised an algorithm to automatically enforce the criteria of
\citet{Ku1977}.
Their algorithm divides the tesseroid into smaller ones instead of increasing
the number of point masses per tesseroid.
A tesseroid is divided if the minimum distance to the computation point
is smaller than the largest dimension of the tesseroid.
This division is repeated recursively until all tesseroids obey the criterion.
Then, GLQ integration is performed for each of the smaller tesseroids
using the specified number of point masses.
The advantage of this adaptive discretization over
increasing the number of points masses is that the
total distribution of point masses will be greater
only close to the computation point.
This makes the adaptive discretization more computationally efficient.

\citet{Grombein2013} developed optimized formula for the gravitational fields
of tesseroids using Cartesian integral kernels.
These formulas are faster to compute and do not have singularities at the poles
like their spherical counterparts.
The Cartesian formulae are numerically integrated using a Taylor series
expansion as per \citet{Heck2007}.
\citet{Grombein2013} use a near-zone separation to mitigate the increased error
at high latitudes.
In the so called ``near-zone'' of the computation point they use a finer
discretization composed by smaller tesseroids.
This is accomplished by dividing the tesseroids along their horizontal
dimensions.
However, the determination of an optimal size of the near-zone remains an
open question \citep{Grombein2013}.

@@

\noindent
where $\alpha, \beta \in \{x, y, z\}$,
$\rho$ is the density,
$G = 6.674\times10^{-11}\ m^3kg^{-1}s^{-1}$ is the gravitational constant,
and

@@

Below, we describe the algorithm to calculate
the effect of a single tesseroid from the input model
on a single computation point.
The algorithm starts by creating an empty stack of tesseroids.
Then, the stack is initialized with the single input tesseroid.
The initialization is done by copying the tesseroid into the stack
and setting \emph{stktop} to zero (the first element).
It is important to note that the stack is not the input tesseroid model.
Instead, it is a buffer used to temporarily store
each stage of the discretization algorithm.

@@

\begin{enumerate}
    \item ``Pop'' the stack (i.e., take the last tesseroid from it).
        This will cause \emph{stktop} to be reduced by one.
        This tesseroid is the one that will be evaluated in the following
        steps.
    \item Compute the distance $d$ (equation~\ref{eq:distance}) between
        the geometric center of the tesseroid and
        the computation point.
    \item Compute the dimensions of the tesseroid $L_\lambda$, $L_\phi$,
        and $L_r$ using equations~\ref{eq:sizelon}-\ref{eq:sizer}.
    \item Check the condition in equation~\ref{eq:condition} for each
        dimension of the tesseroid.
    \item If all dimensions hold the inequality \ref{eq:condition},
        the tesseroid is not divided and its
        gravitational effect is computed using the
        Gauss-Legendre Quadrature
        (equations~\ref{eq:tesspot}-\ref{eq:tesstensor} and~\ref{eq:glq3d}).
        We use a GLQ order of two for all three dimensions
        ($2 \times 2 \times 2 = 8$ point masses)
        by default.
        This value can be changed using
        a command-line argument of the modeling programs.
    \item If any of the dimensions fail the condition:
    \begin{enumerate}
        \item Divide the tesseroid in half along each dimension that failed
             the condition.
        \item Check if there is room in the stack
            for the new tesseroids
            (i.e.,the number of new elements plus \emph{stktop}
             is smaller than the maximum stack size).
             If there isn't,
             warn the user of a ``stack overflow''
             and compute the effect of the tesseroid, as in step 5.
             If there is room in the stack,
             place the smaller tesseroids into
             the stack.
    \end{enumerate}
    \item Repeat the above steps until the stack is empty
        (\emph{stktop} is equal to -1).
\end{enumerate}

@@

The IPython notebooks
and the data generated for the error analysis,
as well as instructions for installing the software
and running the programs,
are also included in
the source code archive that accompanies this article.
Alternatively,
all accompanying material is available
in an online repository\footnote{
https://github.com/pinga-lab/paper-tesseroids}.

@@

For the potential $V$, a distance-size ratio $D=1$ guarantees that the curves
for all experiments are below the $0.1\%$ error threshold.
For $g_z$, the same is achieved with $D=1.5$.
Conversely, $g_{zz}$ requires a value of $D=8$ to achieve an error level of
$0.1\%$.
For a computation height of 260 km, the error curve for $g_{zz}$
intercepts the error threshold line at $D=2.5$.
This behavior suggests that the error curves for $g_{zz}$ might depend on the
computation height.
To test this hypothesis, we computed the error curves for $g_{zz}$ at
heights 2, 10, 50, 150, and 260 km.
Figure~\ref{fig:gzz-with-height} shows the results for $g_{zz}$ at varying
computation heights.
Notice that the distance-size ratio required to achieve $0.1\%$ accuracy
decreases as the computation height increases.
For example, computation at 260 km height requires $D=2.5$ whereas at 10 km
height a value of $D=5.5$ is required to achieve the same accuracy.
One can take advantage of this behavior to reduce the distance-size ratio for
computations of the gravity gradient tensor at high altitudes,
saving computation time.

@@
